A biometric analysis of infant mortality and temperature, northern Sweden 1895-1950

### 2022-01-24 paper.Rmd
major remake, B-P out

---
title: "A biometric analysis of infant mortality and temperature, northern Sweden 1895-1950"
author: "Göran Broström and Tommy Bengtsson"
date: "`r Sys.time()`"
output:
  bookdown::pdf_document2:
    citation_package: natbib
    keep_tex: yes
    number_sections: yes
    toc: yes
    toc_depth: 3
  bookdown::word_document2:
    toc: no
    toc_depth: 2
  bookdown::html_document2:
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: yes
citation_package: natbib
classoption: titlepage
bibliography: bio.bib
titlepage: yes
biblio-style: apalike
documentclass: article
fontsize: 11pt
header-includes:
- \usepackage[utf8]{inputenc}
- \usepackage{graphicx}
- \usepackage{a4wide}
- \usepackage[figuresonly]{endfloat}
- \renewcommand{\efloatseparator}{\mbox{}}
abstract: "The effect of extreme temperatures on infant mortality in the Umeå region 1895-1950 is studied in a biometric analysis setting. More precisely, the effect of climate and weather, measured by temperature, is investigated. It turns out that climate is more important than weather, low average temperatures are more important than temporary dips in temperature. The investigated geographical region is situated too far north for bad effects of high temperatures to show."
---

@@

One important reason for studying neonatal and postneonatal mortality separately is the 
empirical findings by Bourgeois-Pichat [@bp51a;  @bp51b] about endogenous and exogenous mortality and the log-cube transform,
so we start by recollecting his results and formulate them in a modern statistical setting.

# The Bourgeois-Pichat model

Bourgeois-Pichat (*B-P*) suggests that mortality
during the last eleven months of infancy follows a *uniform* distribution
given a specific transformation of time (age), the 
*log-cube transformation* $g$, see equation \@ref(eq:g),

\begin{equation*}\label{eq:g}
g(t) = \log^3(1 + t), \quad 0 \le t \le 365,
\end{equation*}

where $t$ is age measured in *days*. 
Its graph is shown in Figure \@ref(fig:figgraph).

```{r figgraph, fig.cap = "The log-cube transform g.", fig.height = 4}
##g <- function(x) (log(1 + x))^3  # Note: days!
x <- seq(0, 365, length = 1000)
y <- g(x)
plot(x, y, type = "l", xlab = "age (days)", ylab = "g(age)", axes = FALSE, 
     col = "blue")
xat <- c(0, 28, 183, 365)
yat <- g(xat)
axis(1, at = xat)
axis(2, at = yat, labels = round(yat, 1), las = 1)
box()
lines(c(28, 28), c(0, g(28)), lty = 2, col = "darkgreen")
lines(c(0, 28), c(g(28), g(28)), lty = 2, col = "darkgreen")
##
lines(c(183, 183), c(0, g(183)), lty = 2, col = "darkgreen")
lines(c(0, 183), c(g(183), g(183)), lty = 2, col = "darkgreen")
##
lines(c(365, 365), c(0, g(365)), lty = 2, col = "darkgreen")
lines(c(0, 365), c(g(365), g(365)), lty = 2, col = "darkgreen")
abline(h = 0, v = 0)
##abline(a = 0, b = 1, col = "red")
```

The *B-P* model specifies that, on the transformed age scale, the *exogeneous*
mortality distribution, conditional on death before age 365 days,  is *uniform*, that is, the 
survivor function of $T$ is *linear*,

\begin{equation}
S(t) = 1 - \alpha t, \, 0 \le t \le g(365).
\end{equation}

Furthermore, it is assumed that the *endogeneous* component is non-zero only during the first
28 days of life.

Next we apply the model to our infant mortality data from the Umeå-Skellefteå region.

## Infant mortality, a first look

```{r readdata}
infdat <- readRDS("data/infdat3.rds")
```

From the Umeå-Skellefteå region, we follow all persons born in the 
region 1895--1950 until death, reaching the age of 1 year, or leaving follow-up for whatever 
reason, whichever comes first. The first few lines (infants) are shown in Table 
\@ref(tab:infdat3). Note that covariates are left out for the time being.

If `event == TRUE`, then the infant died at age (in days) given by `exit`, otherwise
the infant survived (at least) 365 days and follow-up is lost at `exit == 365`. 
So, in Table \@ref(tab:infdat3), infants No. 1--4 survived, 
while infant No. 5 died at age 162 days. We assume here that all infants are followed from birth,
that is, `enter == 0` in all cases (no left truncation).

```{r infdat3}
source("R/tbl.R")
sinf <- infdat[, c("enter", "exit", "event")]
x <- head(sinf, 4)
y <- sinf[sinf$event == 1, ][1, ]
z <- rbind(x, y)
z$id <- 1:5
z <- z[, c("id", "enter", "exit", "event")]
tbl(z, caption = "A few infants in the data set.")
```

### The Bourgeois-Pichat procedure

The B-P procedure consists of three steps:

1.   Order the data set ascending by `exit`,

2.   apply the log-cube transformation to the age data, and

3.   plot the cumulative sum of the number of deaths vs. age.

The result is shown in Figure \@ref(fig:ggplot2), together with the original
data.  The transform data is stretched  stretched to 365 for easier comparison.

```{r ggplot2, fig.cap = "B-B plots with and without time transform, Umeå-Skellefteå 1895-1950.", fig.height = 4}
nbths <- NROW(infdat)
##oldpar <- par(mfrow = c(2, 1))
sur <- with(infdat, Surv(enter, exit, event))
rs <- risksets(sur)
ss <- cumsum(rs$n.events)
## Natural:
xt <- rs$risktimes
att <- c(0, 28, 183, 365)
plot(xt, ss / nbths, ylim = c(0, max(ss / nbths)), type = "l",
     axes = FALSE, col = "blue", lwd = 2, xlab = "Age (days)",
     ylab = "Cum. deaths")
abline(h = 0)
axis(2, las = 1)
axis(1, at = att, labels = att)
box()
## g-transformed:
lines(g(xt) / g(365) * 365, ss / nbths, 
     col = "red", lwd = 2, xlab = "Age (days)", lty = 2)
##abline(h = 0)
##axis(2, las = 1)
axis(3, at = g(att) / g(365) * 365, labels = att, col = "red", lty = 2)
box()
abline(v = 0)
abline(v = g(c(28, 183, 365)) / g(365) * 365, col = "red", lty = 3)
abline(v = c(28, 183, 365), col = "blue", lty = 3)
##
legend("bottomright", legend = c("Natural", "g-transform"), col = c("blue", "red"), lty = 1)
abline(h = (ss / nbths) [28], lty = 3 )
abline(h = (ss / nbths) [180], lty = 3 )
##par(oldpar)
```

The curves are the empirical *cumulative distribution function (CDF)* expressed 
on the two scales, that is, one minus the *Kaplan-Meier [@km58] estimates* of the 
*survivor functions*.

### Postneonatal mortality

We begin by studying mortality after age 28 days on the $g$-scale, *B-P* predicts that it is approximately *uniform*. 
The uniform distribution, however, is not very useful in a survival analysis context, so we will
investigate a close alternative, the *exponential * distribution. Modifying the *B-P* plot by using successive 
divisions by current *risk set size* instead of original sample size should result in an approximately linear graph if the 
exponential assumption is valid.

```{r pneo, fig.cap = "Relation between cumulative hazards and CDF for transformed postneonatal mortality data."}
ginfdat <- age.window(infdat, c(28, 365))
ginfdat$enter <- g(ginfdat$enter)
ginfdat$exit <- g(ginfdat$exit)
grs <- risksets(with(ginfdat, Surv(enter, exit, event)), members = FALSE)
fit <- coxreg(Surv(enter, exit, event) ~ 1, data = ginfdat)
ser <- plot(fit, fn = "sur", fig = FALSE)
##plot(fit, col = "blue", axes = FALSE)
cum <- fit$hazards[[1]]
x <- cum[, 1]
ch <- cumsum(cum[, 2])
plot(x, ch, type = "s", axes = FALSE, col = "blue", xlab = "Age (days)", ylab = "")
axis(1, at = c(g(28), g(183), g(365)), labels = c(28, 183, 365))
axis(2, las = 1)
abline(h = 0)
box()
x <- ser[[1]][[1]][, 1]
y <- ser[[1]][[1]][, 2]
lines(x, 1 - y, col = "red", lty = 2)
legend("topleft", legend = c("Cumulative hazards", "CDF (Bourgeois-Pichat)"), col = c("blue", "red"),
       lty = c(1, 2), cex = 0.7)
```

As is evident from Figure \@ref(fig:pneo), the difference between the the two curves is
negligible, and both are well approximated by a straight line, so we will from now on work with the 
assumption that the transformed postneonatal data are *exponentially distributed*.
There is also a mathematical argument motivating the similarity: The cdf for an exponentially 
distributed random variable $T$ is

\begin{equation*}
F_T(t) = 1 - \exp(-\lambda t), \quad t, \lambda > 0, 
\end{equation*}

and the first order approximation of this expression is

\begin{equation*}
F_T(t) \approx \lambda t, \quad t, \lambda > 0,
\end{equation*}

which is the *cumulative hazards* function of $T$,
and the approximation is good for small values of $\lambda t$.

So, the *hazard function* (which is the slope of the lines in Figure \@ref(fig:pneo))
is simply estimated by the ratio of the *number of deaths* to the *total time of exposure*
(an *occurrence/exposure* rate):

```{r simpest, echo = TRUE}
posthaz <- with(ginfdat, sum(event) / sum(exit - enter))
```

which evaluates to `r format(posthaz, digits = 3, scientific = TRUE)`. A small number, but remember
that the time unit is defined on the scale defined by the function $g$ (close to *day*).

```{r excum, echo = FALSE}
fitex <- pchreg(Surv(enter, exit, event) ~ 1, cuts = g(c(28, 365)), data = ginfdat)
```

### Neonatal mortality

Neonatal mortality is, according to Bourgeois-Pichat, the sum of two components, *external* and *internal* causes. According to the *B-P*
theory, the external component is equal to the postneonatal mortality, which contains 
only an external component.

We create the neonatal data set by censor all lives at the age 28 days:

```{r neon, echo = TRUE}
neo <- age.window(infdat, c(0, 28))
neo$enter <- g(neo$enter)
neo$exit <- g(neo$exit)
```

```{r defcuts}
neocuts <- c(0, 2, 5, 8, 12, 17, 22, 25, 28)
pcuts <- c(neocuts, 365)
gneocuts <- g(neocuts)
gpcuts <- g(pcuts)
```

and then fit a piecewise constant hazards model with cutpoints at g(`r neocuts`), see 
Figure \@ref(fig:pchneo).

```{r pchneo, fig.cap = "Neonatal hazard function with external component (dashed).", fig.height = 3}
fitneo <- pchreg(Surv(enter, exit, event) ~ 1, data = neo, cuts = gneocuts)
plot(fitneo, fn = "haz", col = "blue", main = "", xlab = "Age (days)", axes = FALSE, lwd = 1.5)
axis(1, at = gneocuts, labels = neocuts)
axis(2, las = 1)
box()
abline(h = 0, v = 0)
abline(h = posthaz, col = "red", lty = 2)
```

Let us take a closer look at what happens at the end of the neonatal interval, see 
Figure \@ref(fig:pchneo2). It shows a smooth transition towards the postneonatal 
(external) mortality.

```{r pchneo2, fig.cap = "Neonatal hazard function with external component (dashed), right hand tail.", fig.height = 3}
##fitneo <- pchreg(Surv(enter, exit, event) ~ 1, data = neo, cuts = neocuts)
par(las = 1)
plot(fitneo, fn = "haz", col = "blue", main = "", xlab = "Age (days)", lwd = 1.5,
     xlim = g(c(5, 28)), axes = FALSE)
axis(1, at = gneocuts, labels = neocuts)
axis(2)
box()
abline(h = 0, v = 0)
abline(h = posthaz, col = "red", lty = 2, lwd = 1.5)
```

### Putting it all together

Looking at the full infant mortality interval on the $g$-scale, we have a piecewise 
constant hazards model with cutpoints at $g(0, 2, 5, 8, 12, 17, 22, 25, 28, 365)$, see 
Figure \@ref(fig:together).

```{r together, fig.cap = "Hazard function, infant mortality, transformed time.", fig.height=6}
ginfdat <- infdat
ginfdat$exit <- g(ginfdat$exit)
fitall <- pchreg(Surv(enter, exit, event) ~ 1, data = ginfdat, cuts = gpcuts)
plot(fitall, fn = "haz", col = "blue", main = "", xlab = "Age (days)", axes = FALSE,
     lwd = 1.5)

axis(1, at = gpcuts, labels = pcuts)
axis(2, las = 1)
box()
abline(h = 0, v = 0)
```

<!--
Now we would like to see the content of Figure \@ref(fig:together) on the natural time 
scale. If $Y = g(T)$, we have the relation between their respective hazard functions

\begin{equation*}
h_T(t) = \frac{d}{dt} g(t) h_Y\bigl(g^{-1}(t)\bigr), \quad 0 \le t \le 365,
\end{equation*}

and with $h_Y$ shown in Figure \@ref(fig:together), we get $h_T$ in Figure 
\@ref(fig:allnorm) **(must be checked!)**.

```{r allnorm, fig.cap = "Hazard function, infant mortality, natural time scale.", eval = FALSE}
y <- hazards(fitall, cum = FALSE, n.points = 3000)$x
hy <- hazards(fitall, cum = FALSE, n.points = 3000)$y
te <- ginv(y)
hte <- dg(te) * hy
plot(te, hte, type = "s", col = "blue", xlab = "Days", ylab = "", axes = FALSE,
     ##ylim = c(0, 0.002)) 
     xlim = c(0, 28))
axis(1, at = pcuts)
axis(2, las = 1)
box()
abline(h = 0, v = 0)
```

-->

The way to disentangle the endogenous and exogenous components is to start with a plot 
of the *cumulative* hazard function and add the exogenous 
part from the postneonatal estimate, see Figure \@ref(fig:separ).

```{r separ, fig.cap = "Separation of exogenous and endogenous infant mortality (cumulative hazards, per cent).", fig.width = 5, fig.height = 4}
plot(fitall, fn = "cum", col = "blue", main = "", xlab = "Age (days)", axes = FALSE,
     lwd = 1.5, ylab = "Cumulative hazards")

axis(1, at = c(0, g(28), g(365)), labels = c(0, 28, 365), pos = 0)
axis(2, las = 1, pos = 0, at = c(0, 0.0096, 0.0338), labels = c(0, 0.96, 3.38))
#box()
abline(v = 0)
texhaz <- g(365) * fitex$hazards[1, 1]
lines(c(0, g(365)), c(0, texhaz), col = "red", lwd = 1.5)
atte <- c(0.0000, 0.0763, texhaz)
axis(4, at = atte, labels = 100 * round(atte, 4), las = 1, cex = 0.5,
     pos = g(365))
abline(v = g(365))
text(100, 0.03, "Exogenous", col = "red", srt = 22, cex = 0.8)
text(100, 0.053, "Endogenous plus exogenous", col = "blue", srt = 22, cex = 0.8)
axis(3, at =  c(0, g(28), g(365)), labels = c(0, 28, 365))
lines(c(0, g(28)), c(0.0096, 0.0096), lty = 3)
lines(c(0, g(28)), c(0.0338, 0.0338), lty = 3)
abline(v = g(28))
```

So we can see from Figure \@ref(fig:separ) (vertical axis on the right hand side) that of the total mortality, 
`r round(5.27 / 7.63 * 100)` per cent is exogenous and the rest is endogenous.
Similarly, of the neonatal mortality, `r round(0.96 / 3.38 * 100)` per cent is
exogenous and the rest is endogenous (vertical axis on the left hand side).
As always, this requires that we trust the Bourgeois-Pichat model.

# Infant mortality, climate and weather

## Data

### Infant mortality, summary graphs

@@

### Temperature

@@

Temperature data are collected from two weather stations, *Umeå* and *Bjuröklubb*
(used with population data from Skellefteå coastal area). Both stations deliver
daily temperature data covering our time period, usually three measures per day, morning,
noon, and evening. Weekly averages are calculated year by year, and deviations from 
the averages of the weekly averages are used as a time-varying *communal covariate*. See Figure 
\@ref(fig:monavgtemp) for the monthly distributions.

@@

## Statistical modelling

@@

## Postneonatal mortality

@@

## Neonatal mortality

@@

### Girls

@@

### Boys

@@

## Postneonatal mortality, winter and spring

@@

## Neonatal mortality, winter and spring

>>>>

---
title: "A biometric analysis of infant mortality and temperature, northern Sweden 1895-1950"
author: "Göran Broström and Tommy Bengtsson"
date: "`r Sys.time()`"
output:
  bookdown::pdf_document2:
    citation_package: natbib
    keep_tex: yes
    number_sections: yes
    toc: yes
    toc_depth: 3
  bookdown::word_document2:
    toc: no
    toc_depth: 2
  bookdown::html_document2:
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: yes
citation_package: natbib
classoption: titlepage
bibliography: bio.bib
titlepage: yes
biblio-style: apalike
documentclass: article
fontsize: 11pt
header-includes:
- \usepackage[utf8]{inputenc}
- \usepackage{graphicx}
- \usepackage{a4wide}
- \usepackage[figuresonly]{endfloat}
- \renewcommand{\efloatseparator}{\mbox{}}
abstract: "The effect of extreme temperatures on infant mortality in the Umeå region 1895-1950 is studied in a biometric analysis setting. More precisely, the effect of climate and weather, measured by temperature, on infant mortality is investigated. It turns out that climate is more important than weather, low average temperatures are more important than temporary dips in temperature. The investigated geographical region is situated too far north for bad effects of high temperatures to show."
---

@@

One important reason for studying neonatal and postneonatal mortality separately is the 
empirical findings by Bourgeois-Pichat [@bp51a;  @bp51b] about endogenous and exogenous 
mortality and the log-cube transform.

# Data

We have two sources of data which we combine into one data set suitable for our purpose.
The first is demographic data obtained from the 
*Centre for Demographic and Ageing Research* (CEDAR, https://cedar.umu.se),
the second is daily temperature measurements obtained from the 
*Swedish Meteorological and Hydrological Institute* (SMHI, https://www.smhi.se).

## Infant mortality

Individual data with all births between 1 January 1895 and 31 December 1950 in two 
coastal areas of north Sweden, Skellefteå (51560 births) and Umeå (31213 births).
They were followed until death or age 365 days, whichever came first. 
The following *static* characteristics were observed on each child:

**birthdate** Date of birth.

**sex** Girl or boy.

**exit** Number of days under observation.

**event** Logical, *TRUE* if a death is observed.

**socBranch** Working branch of father (if any).

**socStatus** Social status of family, based on HISCLASS.

**illeg** Mother unmarried?

**parity** Order among siblings.

Some crude statistics about infant, neonatal, and postneonatal mortality are shown in 
Figures.

@@

## Temperature

@@

Temperature data are collected from two weather stations, *Umeå* and *Bjuröklubb*
(used with population data from Skellefteå coastal area). Both stations deliver
daily temperature data covering our time period, usually three measures per day, morning,
noon, and evening. In Table \@ref(tab:typweek), the Umeå data from the week 1--7 January,
1923 is shown.

```{r typweek}
tum <- read.table("~/Forskning/Data/ume_temp.csv", skip = 10, sep = ";", header = TRUE)
source("R/tbl.R")
ww <- tum[70137:(70137 + 20), ]
names(ww) <- c("Date", "Time", "Temperature", "Quality")
tbl(ww, caption = "Raw temperature data from first week of 1923, Umeå.", fs = 10,
    linesep = c("", "", "\\addlinespace"))
```

There are three measurements per day, or 21 per week. In the forthcoming analyses,
the weekly data are summarized in a few measurements, see Table \@ref(tab:cond).

```{r cond}
sh <- umetemp[umetemp$year == 1923 & umetemp$week == 1, ]
tbl(round(sh[, 1:8], 2), caption = "Weekly summarized temperature data: Umeå 1923, first week.", fs = 9)
```

Weekly averages are calculated year by year, and deviations from 
the averages of the weekly averages are used as a time-varying *communal covariate*. See Figure 
\@ref(fig:monavgtemp) for the monthly distributions.

@@

## Mortality and temperature as a communal covariate

The two data sets, mortality and weather, are combined ino one by treating temperature data 
as a communal covariate and incorporate it as such in the mortality data set.

# Statistical modelling

@@

# Postneonatal mortality

@@

# Neonatal mortality

@@

## Girls

@@

## Boys

@@

# Postneonatal mortality, winter and spring

@@

# Neonatal mortality, winter and spring
